Automating Free Logic in HOL, with an Application in Category Theory\thanks{Mention DFG grants ...}

### 2017-03-30 FreeLogic/2017-Journal/article.tex
update

\newcommand\hol[1]{\boldsymbol{#1}}
\newcommand\lift[1]{\lceil #1 \rceil}
\newcommand\llift[1]{\dot{#1}}

@@

The purpose of this article is not to claim any novel results in category
theory, but to demonstrate an elegant way to ``implement'' and utilise
reasoning in free logic, and to present respective experiments.

@@

\section{Preliminaries}
\label{sec:preliminaries}
\subsection{Free Logic}
Free logic \cite{Lambert60,Scott67,lambert02:_free_logic} refers to a class of logic
formalisms that are free of basic existence assumptions regarding the
denotation of terms. Remember that terms in e.g. traditional classical
and intuitionistic predicate logics always denote an (existing) object
in a given (non-empty) domain $\D$, and that $\D$ is also exactly the set the
quantifiers range over. In free logic these
basic assumption are abolished. Terms do still denote objects in a
(non-empty) domain $\D$, but a (possibly empty) set $\ex\subseteq \D$ is
chosen to characterize the subdomain of ``existing'' resp. ``defined''
objects in $\D$. Quantification is
now restricted to set $\ex$ of existing/defined objects only. It is
obvious how this can be used to model undefideness and partiality:
problematic terms, e.g. division by zero or improper definite
descriptions, still denote, but they refer to undefined objects, that
is, objects $d$ in $\D\setminus \ex$ lying outside of the scope of
quantification.

@@

\begin{definition}[Dual-domain semantics of $\FFOL$]
  A \emph{model structure for $\FFOL$} consists of a triple
  $\langle D, \ex, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $\ex\subseteq D$ a possible empty set of ``existing''
  resp. ``defined'' objects, and $I$ an interpretation function
  mapping 0-ary function symbols (constants) to defined objects $d\in \ex$, 
 0-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \ldots \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \ldots \times D$. Finally, $\star\in D\setminus \ex$ is a designated
  (non-existing/undefined) object.

  Given a variable assignment $g$, we
  define the interpretation function $\|\cdot\|^{I,g}$ for terms and formulas of
  $\FFOL$ as follows:
\noindent
\begin{description}
\item{Terms}
\begin{enumerate}
\item $\| x \|^{I,g} = g(x)$ for variable symbols $x\in\V$
\item $\| c \|^{I,g} = I(c)$, where $c\in\F$ is an 0-ary function symbol
\item $\| f(t_1,\ldots,t_n)\|^{I,g} = I(f)(\| t_1 \|^{I,g},\ldots,\|
  t_n \|^{I,g})$, where $f\in\F$ is an $n$-ary ($n\geq 1$) function symbol
\item $\| \Descr x. r \|^{I,g} = d \in \ex$, such that $\| r
  \|^{I,g[d/x]} = \True$ and  $\| r
  \|^{I,g[d'/x]} = \False$ for all $d'\not=d\in \ex$ (i.e. $d$ is the
  unique existing object for which $r$ holds); if there is no
  such $d\in \ex$, then $\| \Descr x. r \|^{I,g} = \star$
\end{enumerate}

\item{Formulas}
\begin{enumerate}
\item $\| p \|^{I,g} = I(p)$, where $p\in\F$ is an 0-ary predicate
  symbol
\item $\| t_1 = t_2 \|^{I,g} = \True$ iff $\| t_1 \|^{I,g} = \| t_2
  \|^{I,g}$ (this basic notion of primitive equality on $D$ implies that equations such as $1/0 = 1/0$ are
  evaluated to $\True$; later, in Section \ref{??}, we will
  define and utilise further notions of equality, including \emph{Kleene equality}
  and \emph{existing equality}).
\item $\| p(t_1,\ldots,t_n) \|^{I,g} = \True$ iff $(\| t_1 \|^{I,g},\ldots,\|
  t_n \|^{I,g}) \in I(p)$ for $n$-ary ($n\geq 1$) predicate symbols
  $p\in\P$
\item $\| \neg r \|^{I,g} = \True$ iff $\| r \|^{I,g} = \False$
\item $\| r \Impl s \|^{I,g} = \True$ iff $\| r \|^{I,g} = \False$
  or $\| s \|^{I,g} = \True$
\item $\| \forall x. r  \|^{I,g} = \True$ iff for all $d \in \ex$ we
  have $\|  r  \|^{I,g[d/x]} = \True$ % (here $g[d/x]$ denotes the
  % variable assignment that is identical to $g$, except for variable
  % $x$, which is now mapped to object $d$)
\end{enumerate}
\end{description}

@@

\subsection{Classical Higher-Order Logic}
Short introduction ... todo.

\begin{definition}\label{holtypes}
  The set $\hol{T}$ of simple types freely generated from a set of
  basic types $\{\hol{o}, \hol{\mu}, \hol{\iota}\}$ using the function type
  constructor $\hol{\ar}$. $\hol{o}$ is the type of Booleans,
  $\hol{\mu}$ is the type of individuals, and type
  $\hol{\iota}$ is employed as the type of possible worlds below. We
  may avoid parentheses if the structure of a complex type is clear in context.
\end{definition}

\begin{definition}\label{holgrammar}
The language of higher-order logic $\HOL$ is defined by the following
grammar:\footnote{Ednote: Maybe better to define this analagous to $\FFOL$.}
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\, t_\alpha)_\beta} \mid  \hol{\neg_{o\ar o}\, s_o} \mid \\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} 
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. $p_\alpha$ denotes typed constants and
$X_\alpha$ typed variables (distinct from $p_\alpha$).  Complex typed
terms are constructed via abstraction and application. The type of
each term is given as a subscript.  Terms $s_o$ of type $o$ are called
formulas.  The logical connectives of choice are
$\hol{\neg_{o\ar o}}$, $\hol{\vee_{o\ar
  o\ar o}}$, and $\hol{\forall_{(\alpha\ar
  o)\ar o}}$ (for $\hol{\alpha\in T}$).  Type subscripts may be dropped if
irrelevant or obvious. Similarly, parentheses may be avoided.  Binder
notation $\hol{\all{X_\alpha} s_o}$ is used as shorthand for
$\hol{\forall_{(\alpha\ar o)\ar o}
(\lam{X_\alpha} s_o)}$, and infix notation $\hol{s \vee t}$ is employed
instead of $\hol{((\vee s)\, t)}$. From the above connectives, other logical
connectives, such as $\hol{\top}$, $\hol{\bot}$, $\hol{\wedge}$,
$\hol{\supset}$, $\hol{\equiv}$ and $\hol{\exists}$, can be defined in the usual way.
\end{definition}

@@

%\begin{definition}\label{homlbetaeta} %\footnote{\textbf{Chris:} Is Def.4 needed at all?}
  Two common relations on terms are given by $\hol{\beta}$-reduction and
  $\hol{\eta}$-reduction.  A $\hol{\beta}$-redex has the form $\hol{(\lam{X}s)t}$ and
  $\hol{\beta}$-reduces to $\hol{[t/X]s}$.  An $\hol{\eta}$-redex has the form
  $\hol{(\lam{X}s X)}$ where variable $\hol{X}$ is not free in $\hol{s}$; it
  $\hol{\eta}$-reduces to $\hol{s}$.  We write $\hol{s\eqb t}$ to mean $\hol{s}$ can be
  converted to $\hol{t}$ by a series of $\hol{\beta}$-reductions and expansions.
  Similarly, $\hol{s\eqbe t}$ means $\hol{s}$ can be converted to $\hol{t}$ using both
  $\hol{\beta}$ and $\hol{\eta}$.  For each $\hol{s_\alpha\in \HOL}$ there is a unique
  \emph{$\beta$-normal form} and a unique \emph{$\beta\eta$-normal
    form}.
% From this fact we know $s\eqb t$ ($s\eqbe t$) iff
% $\Bnormform s\Metaeq\Bnormform t$
% ($\Benormform s\Metaeq\Benormform t$).
% The semantics of $\stt$ is well understood and thoroughly documented
% in the literature \cite{Henkin50,Andrews72b,Andrews72a,BBK04}; our
% summary below is adapted from Andrews~\cite{sep-type-theory-church}.
%\end{definition}

%\begin{definition}\label{homlassignment}
A \emph{variable assignment} $g$ maps
variables $\hol{X_\alpha}$ to elements in $\hol{D_\alpha}$. $\hol{g[d/W]}$ denotes the
assignment that is identical to $\hol{g}$, except for variable $\hol{W}$, which is
now mapped to $\hol{d}$.
%\end{definition}

@@

\begin{definition}\label{holvalue}
  The \emph{value} $\hol{\| s_\alpha\|^{M,g}}$ of a \HOL term
  $\hol{s_\alpha}$ on a model $\hol{M}=\hol{\langle D, I \rangle}$ under assignment $\hol{g}$ is an element $\hol{d}\in \hol{D_\alpha}$
  defined in the following way:
\begin{enumerate}
\item $\hol{\|p_\alpha\|^{M,g}} = \hol{I(p_\alpha)}$ and $\hol{\|X_\alpha\|^{M,g}} = \hol{g(X_\alpha)}$
\item $\hol{\|(s_{\alpha\ar\beta}\, t_\alpha)_\beta\|^{M,g}} = \hol{\|s_{\alpha\ar\beta}\|^{M,g}(\|t_\alpha\|^{M,g})}$
\item $\hol{\|(\lam{X_\alpha}s_\beta)_{\alpha\ar\beta}\|^{M,g}} = $
  the function $\hol{f}$ from $\hol{D_\alpha}$ to $\hol{D_\beta}$ such
  that $\hol{f(d)} = \hol{\|s_\beta\|^{M,g[d/X_\alpha]}}$ for all
  $\hol{d}\in \hol{D_\alpha}$
\item $\hol{\|(\neg_{o\ar o}\, s_o)_o\|^{M,g}} = \hol{T}$ iff $\hol{\|s_o\|^{M,g}} = \hol{F}$
\item $\hol{\|(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)_o\|^{M,g}} =
  \hol{T}$ iff\, $\hol{\|s_o\|^{M,g}} = \hol{T}$ or $\hol{\|t_o\|^{M,g}}
  = \hol{T}$
\item $\hol{\|(\forall_{(\alpha\ar o)\ar o}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{T}$ iff for all $\hol{d}\in
  \hol{D_\alpha}$ we have $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = \hol{T}$
%\item $\|(\Box_{o\ar o}\, s_o)_o\|^{M,g,w} = T$ iff for all $v \in W$ with $w R v$ we have  $\|s_o\|^{M,g,v}  = T$
\end{enumerate}
\end{definition}

\begin{definition}\label{homlhenkinmodel}
 A model $\hol{M}=\hol{\langle D, I \rangle}$ is called a
 \emph{standard model} iff for all $\alpha,\beta\in T$ we have
 $D_{\alpha\ar\beta} = \{ f \mid f : D_\alpha \longrightarrow D_\beta
 \}$. In a \emph{Henkin model} function spaces are not necessarily
 full. Instead it is only required that $D_{\alpha\ar\beta}
 \subseteq \{ f \mid f : D_\alpha \longrightarrow D_\beta \}$ (for all
 $\alpha,\beta\in T$) and that the valuation function 
 $\|\cdot\|^{M,g}$ from above is total (i.e., every term denotes). Any standard model is obviously
 also a Henkin model. We consider Henkin models in the remainder.
\end{definition}

\begin{definition}\label{homlvalid}
A formula $s_o$ is \emph{true} in model $M$ 
under assignment $g$ iff $\|s_o\|^{M,g} = T$; this is also denoted
as $M,g \models^\HOL s_o$.  A formula $s_o$ is called \emph{valid} in
$M$, which is denoted as $M \models^\HOL s_o$, iff $M,g \models^\HOL s_o$ for all
assignments $g$. Finally, a formula $s_o$ is called
\emph{valid}, which we denote by $\models^\HOL s_o$, iff $s_o$ is valid for
all $M$. Moreover, we write $\Gamma\models^\HOL \Delta$ (for sets of
 formulas $\Gamma$ and $\Delta$) iff there is a model $M$ and an assignment $g$ such that  
 $M,g \models^\HOL s_o$ for all $s_o\in \Gamma$ and $M,g \models^\HOL t_o$ for at least one $t_o\in \Delta$.
\end{definition}

@@

\section{Sound and Complete Embedding of $\FFOL$ in HOL}
Todo ... the embedding in Isabelle/HOL is depicted in 
figure \ref{pic1}

\begin{figure}[t]
 \includegraphics[width=\textwidth]{EmbeddingPic}
 \caption{Isabelle/HOL formalisation of $\FFOL$ in $\HOL$ \label{pic1}}
\end{figure}

In this framework partial and total functions are modelled as follows: 
A function $f$ is total if and only if for all $x$ we have $\ex x \imp \ex(f x)$ 
For partial functions $f$ we may have some $x$ such that $\ex x$ but not
 $\ex(f x)$. A function $f$ is strict  if  and only if for all $x$
we have $\ex(f x) \imp \ex x$.

@@

The stepwise evolution has been described in
\cite{R58}. Below we summarise these experiments.
However, first we describe some basic modeling decisions for the
technical encoding in Isabelle/HOL. The Isabelle/HOL sources of our experiments are
available at \url{todo-sources}.

>>>>

\newcommand\map[1]{\hat{#1}}
\newcommand\hol[1]{\boldsymbol{#1}}
\newcommand\lift[1]{\lceil #1 \rceil}
\newcommand\llift[1]{\dot{#1}}

@@

  The purpose of this article is not to claim any novel results in
  category theory, but to demonstrate an elegant way to ``implement''
  and utilise interactive and automated reasoning in free logic, and
  to present respective experiments.

@@

\section{Preliminaries}
\label{sec:preliminaries}
\subsection{Free Logic}
Free logic \cite{Lambert60,Scott67,lambert02:_free_logic} refers to a class of logic
formalisms that are free of basic existence assumptions regarding the
denotation of terms. Remember that terms in e.g. traditional classical
and intuitionistic predicate logics always denote an (existing) object
in a given (non-empty) domain $\D$, and that $\D$ is also exactly the set the
quantifiers range over. In free logic these
basic assumption are abolished. Terms do still denote objects in a
(non-empty) domain $\D$, but a (possibly empty) set $\ex\subseteq \D$ is
chosen to characterize the subdomain of ``existing'' resp. ``defined''
objects in $\D$. Quantification is
now restricted to set $\ex$ of existing/defined objects only.

It is obvious how this can be used to model undefideness and
partiality: problematic terms, e.g. division by zero or improper
definite descriptions, still denote, but they refer to undefined
objects, that is, objects $d$ in $\D\setminus \ex$ lying outside of
the scope of quantification.
%In this framework partial and total functions are modelled as follows: 
Moreover, a function $f$ is \emph{total} if and only if for all $x$ we have $\ex x \imp \ex(f x)$ 
For \emph{partial} functions $f$ we may have some $x$ such that $\ex x$ but not
$\ex(f x)$. A function $f$ is \emph{strict}  if  and only if for all $x$
 we have $\ex(f x) \imp \ex x$.

@@

\begin{definition}[Dual-domain semantics of $\FFOL$]
  A \emph{model structure for $\FFOL$} consists of a triple
  $\langle D, \ex, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $\ex\subseteq D$ a possible empty set of ``existing''
  resp. ``defined'' objects, and $I$ an interpretation function
  mapping $0$-ary function symbols (constants) to defined objects $d\in \ex$, 
 $0$-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \ldots \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \ldots \times D$. Finally, $\star\in D\setminus \ex$ is a designated
  (non-existing/undefined) object.

  Given a variable assignment $g$, we
  define the interpretation function $\|\cdot\|^{I,g}$ for terms and formulas of
  $\FFOL$ as follows:
\noindent
\begin{description}
\item{Terms}
\begin{enumerate}
\item $\| x \|^{I,g} = g(x)$ for variable symbols $x\in\V$
\item $\| c \|^{I,g} = I(c)$, where $c\in\F$ is an $0$-ary function symbol
\item $\| f(t_1,\ldots,t_n)\|^{I,g} = I(f)(\| t_1 \|^{I,g},\ldots,\|
  t_n \|^{I,g})$, where $f\in\F$ is an $n$-ary ($n\geq 1$) function symbol
\item $\| \Descr x. r \|^{I,g} = d \in \ex$, such that $\| r
  \|^{I,g[d/x]} = \True$ and  $\| r
  \|^{I,g[d'/x]} = \False$ for all $d'\not=d\in \ex$ (i.e. $d$ is the
  unique existing object for which $r$ holds); if there is no
  such $d\in \ex$, then $\| \Descr x. r \|^{I,g} = \star$
\end{enumerate}

\item{Formulas}
\begin{enumerate}
\item $\| q \|^{I,g} = I(q)$, where $q\in\P$ is an $0$-ary predicate
  symbol
\item $\| t_1 = t_2 \|^{I,g} = \True$ iff $\| t_1 \|^{I,g} = \| t_2
  \|^{I,g}$ (this basic notion of primitive equality on $D$ implies that equations such as $1/0 = 1/0$ are
  evaluated to $\True$; later, in Section \ref{??}, we will
  define and utilise further notions of equality, including \emph{Kleene equality}
  and \emph{existing equality}).
\item $\| p(t_1,\ldots,t_n) \|^{I,g} = \True$ iff $(\| t_1 \|^{I,g},\ldots,\|
  t_n \|^{I,g}) \in I(p)$ for $n$-ary ($n\geq 1$) predicate symbols
  $p\in\P$
\item $\| \neg r \|^{I,g} = \True$ iff $\| r \|^{I,g} = \False$
\item $\| r \Impl s \|^{I,g} = \True$ iff $\| r \|^{I,g} = \False$
  or $\| s \|^{I,g} = \True$
\item $\| \forall x. r  \|^{I,g} = \True$ iff for all $d \in \ex$ we
  have $\|  r  \|^{I,g[d/x]} = \True$ % (here $g[d/x]$ denotes the
  % variable assignment that is identical to $g$, except for variable
  % $x$, which is now mapped to object $d$)
\end{enumerate}
\end{description}

@@

\subsection{Classical Higher-Order Logic}
Simple type theory, also referred to as classical higher-order logic
(HOL), is an expressive logic formalism that allows for higher-order
quantification, that is quantification over arbitrary set and function
variables.  It is based on the simply typed
$\lambda$-calculus\footnote{ For a detailed discussion of typed
  $\lambda$-calculi, we refer to the
  literature~\cite{DBLP:books/daglib/0032840}.  }
  %\footnote{
    %Although \sys~actually supports a polymorphic typed $\lambda$-calculus
    %we will
    %stick here to the simply typed $\lambda$-calculus for a better understanding.}
  and it has its origin in the work by
  Church~\cite{DBLP:journals/jsyml/Church40}.

\begin{definition}[Types]\label{holtypes}
  The set $\hol{T}$ of simple types freely generated from a set of
  basic types $\{\hol{o}, \hol{\mu}, \hol{\iota}\}$ using the function type
  constructor $\hol{\ar}$. $\hol{o}$ is the type of Booleans,
  $\hol{\mu}$ is the type of individuals, and type
  $\hol{\iota}$ is employed as the type of possible worlds below. We
  may avoid parentheses if the structure of a complex type is clear in context.
\end{definition}

\begin{definition}[Syntax of $\HOL$]\label{holgrammar}
The language of higher-order logic $\HOL$ is defined by the following
grammar:\footnote{Ednote: Maybe better to define this analagous to $\FFOL$.}
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\, t_\alpha)_\beta} \mid  \hol{\neg_{o\ar o}\, s_o} \mid \\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} \mid \hol{\Descr(\lam{X_\alpha} s_o)}
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. $\hol{p_\alpha}$ denotes typed constants and
$\hol{X_\alpha}$ typed variables (distinct from $\hol{p_\alpha}$).  Complex typed
terms are constructed via abstraction and application. The type of
each term is given as a subscript.  Terms $\hol{s_o}$ of type $\hol{o}$ are called
formulas.  The logical connectives of choice are
$\hol{\neg_{o\ar o}}$, $\hol{\vee_{o\ar
  o\ar o}}$, $\hol{\forall_{(\alpha\ar
  o)\ar o}}$ (for $\hol{\alpha\in T}$) and $\hol{\Descr_{(\alpha\ar
  o)\ar \alpha}}$ (for $\hol{\alpha\in T}$).  Type subscripts may be dropped if
irrelevant or obvious. Similarly, parentheses may be avoided.  Binder
notation $\hol{\all{X_\alpha} s_o}$ and $\hol{\Descr {X_\alpha} s_o}$ is used as shorthand for
$\hol{\forall_{(\alpha\ar o)\ar o}
(\lam{X_\alpha} s_o)}$ and $\hol{\Descr(\lam{X_\alpha} s_o)}$, and infix notation $\hol{s \vee t}$ is employed
instead of $\hol{((\vee s)\, t)}$. From the above connectives, other logical
connectives, such as $\hol{\top}$, $\hol{\bot}$, $\hol{\wedge}$,
$\hol{\Impl}$, $\hol{\equiv}$ and $\hol{\exists}$, can be defined in the usual way.
\end{definition}

@@

%\begin{definition}\label{homlbetaeta} %\footnote{\textbf{Chris:} Is Def.4 needed at all?}
  Two common relations on terms are given by $\hol{\beta}$-reduction and
  $\hol{\eta}$-reduction.  A $\hol{\beta}$-redex has the form $\hol{(\lam{X}s)t}$ and
  $\hol{\beta}$-reduces to $\hol{[t/X]s}$.  An $\hol{\eta}$-redex has the form
  $\hol{(\lam{X}s X)}$ where variable $\hol{X}$ is not free in $\hol{s}$; it
  $\hol{\eta}$-reduces to $\hol{s}$.  We write $\hol{s\eqb t}$ to mean $\hol{s}$ can be
  converted to $\hol{t}$ by a series of $\hol{\beta}$-reductions and expansions.
  Similarly, $\hol{s\eqbe t}$ means $\hol{s}$ can be converted to $\hol{t}$ using both
  $\hol{\beta}$ and $\hol{\eta}$.  For each $\hol{s_\alpha\in \HOL}$ there is a unique
  \emph{$\hol{\beta}$-normal form} and a unique \emph{$\hol{\beta\eta}$-normal
    form}.
% From this fact we know $s\eqb t$ ($s\eqbe t$) iff
% $\Bnormform s\Metaeq\Bnormform t$
% ($\Benormform s\Metaeq\Benormform t$).
% The semantics of $\stt$ is well understood and thoroughly documented
% in the literature \cite{Henkin50,Andrews72b,Andrews72a,BBK04}; our
% summary below is adapted from Andrews~\cite{sep-type-theory-church}.
%\end{definition}

%\begin{definition}\label{homlassignment}
A \emph{variable assignment} $\hol{g}$ maps
variables $\hol{X_\alpha}$ to elements in $\hol{D_\alpha}$. $\hol{g[d/W]}$ denotes the
assignment that is identical to $\hol{g}$, except for variable $\hol{W}$, which is
now mapped to $\hol{d}$.
%\end{definition}

@@

\begin{definition}\label{holvalue}
  The \emph{value} $\hol{\| s_\alpha\|^{M,g}}$ of a \HOL term
  $\hol{s_\alpha}$ on a model $\hol{M}=\hol{\langle D, I \rangle}$ under assignment $\hol{g}$ is an element $\hol{d}\in \hol{D_\alpha}$
  defined in the following way:
\begin{enumerate}
\item $\hol{\|p_\alpha\|^{M,g}} = \hol{I(p_\alpha)}$ and $\hol{\|X_\alpha\|^{M,g}} = \hol{g(X_\alpha)}$
\item $\hol{\|(s_{\alpha\ar\beta}\, t_\alpha)_\beta\|^{M,g}} = \hol{\|s_{\alpha\ar\beta}\|^{M,g}(\|t_\alpha\|^{M,g})}$
\item $\hol{\|(\lam{X_\alpha}s_\beta)_{\alpha\ar\beta}\|^{M,g}} = $
  the function $\hol{f}$ from $\hol{D_\alpha}$ to $\hol{D_\beta}$ such
  that $\hol{f(d)} = \hol{\|s_\beta\|^{M,g[d/X_\alpha]}}$ for all
  $\hol{d}\in \hol{D_\alpha}$
\item $\hol{\|(\neg_{o\ar o}\, s_o)_o\|^{M,g}} = \hol{T}$ iff $\hol{\|s_o\|^{M,g}} = \hol{F}$
\item $\hol{\|(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)_o\|^{M,g}} =
  \hol{T}$ iff\, $\hol{\|s_o\|^{M,g}} = \hol{T}$ or $\hol{\|t_o\|^{M,g}}
  = \hol{T}$
\item $\hol{\|(\forall_{(\alpha\ar o)\ar o}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{T}$ iff for all $\hol{d}\in
  \hol{D_\alpha}$ we have $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = \hol{T}$
\item $\hol{\|(\Descr_{(\alpha\ar o)\ar \alpha}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{d}$ if there exists a unique $\hol{d}\in \hol{D_{\alpha}}$ such that
  $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = \hol{T}$, 
otherwise  $\hol{\|(\Descr_{(\alpha\ar o)\ar \alpha}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{e}$ for an arbitrary element $\hol{e}\in \hol{D_{\alpha}}$
%\item $\|(\Box_{o\ar o}\, s_o)_o\|^{M,g,w} = T$ iff for all $v \in W$ with $w R v$ we have  $\|s_o\|^{M,g,v}  = T$
\end{enumerate}
\end{definition}

\begin{definition}\label{homlhenkinmodel}
 A model $\hol{M}=\hol{\langle D, I \rangle}$ is called a
 \emph{standard model} iff for all $\hol{\alpha,\beta}\in \hol{T}$ we have
 $\hol{D_{\alpha\ar\beta}} = \{ f \mid f : \hol{D_\alpha} \longrightarrow \hol{D_\beta}\}$. In a \emph{Henkin model} function spaces are not necessarily
 full. Instead it is only required that $\hol{D_{\alpha\ar\beta}}
 \subseteq \{ f \mid f : \hol{D_\alpha} \longrightarrow \hol{D_\beta}\}$ (for all
 $\hol{\alpha,\beta}\in \hol{T}$) and that the valuation function 
 $\hol{\|\cdot\|^{M,g}}$ from above is total (i.e., every term denotes). Any standard model is obviously
 also a Henkin model. We consider Henkin models in the remainder.
\end{definition}

\begin{definition}\label{homlvalid}
A formula $\hol{s_o}$ is \emph{true} in model $M$ 
under assignment $\hol{g}$ iff $\hol{\|s_o\|^{M,g}} = \hol{T}$; this is also denoted
as $\hol{M,g \models^\HOL s_o}$.  A formula $\hol{s_o}$ is called \emph{valid} in
$\hol{M}$, which is denoted as $\hol{M \models^\HOL s_o}$, iff $\hol{M,g \models^\HOL s_o}$ for all
assignments $\hol{g}$. Finally, a formula $\hol{s_o}$ is called
\emph{valid}, which we denote by $\hol{\models^\HOL s_o}$, iff $\hol{s_o}$ is valid for
all $\hol{M}$. Moreover, we write $\hol{\Gamma\models^\HOL \Delta}$ (for sets of
 formulas $\hol{\Gamma}$ and $\hol{\Delta}$) iff there is a model $\hol{M}$ and an assignment $\hol{g}$ such that  
 $\hol{M,g \models^\HOL s_o}$ for all $\hol{s_o}\in \hol{\Gamma}$ and $\hol{M,g \models^\HOL t_o}$ for at least one $\hol{t_o}\in \hol{\Delta}$.
\end{definition}

@@

\begin{figure}[tp]
 \includegraphics[width=\textwidth]{EmbeddingPic}
 \caption{Isabelle/HOL formalisation of $\FFOL$ in $\HOL$ \label{pic1}}
\end{figure}

\section{Shallow Semantical Embedding of $\FFOL$ in HOL}

The existence predicate $E$ is realized in the embedding as a $\HOL$
predicate $\hol{\ex}$ of type $\hol{ i\ar o}$; we assume that a
respective constant symbol is given in the signature of \HOL.  The raw domain $\hol{D}$ is simply
associated with the set of all objects of type $\hol{i}$, respectively
with the universal
predicate $\hol{\lambda X_i. T}$. Moreover, let
constant symbol $\hol{\star}$ of type $i$ be in the signature of
\HOL. We assume that $\hol{\|E \star\|^{M,g}=F}$ for all $\hol{M,g}$
i.e. that the element denoted by $\hol{\star}$ is not an element of
the domain of existing objects denoted by $\hol{\ex}$.

\begin{definition}[Embedding of $\FFOL$ in \HOL]
Given a formula $s \in \FFOL$. We map each language construct $s$ in the
signature of $\FFOL$ to a corresponding term $\map{s}$ of \HOL. This
mapping is defined as follows:

\begin{tabular}{lcll}
$\map{x}$    & := & $\hol{X_i}$ &for all $x\in \V$\\
$\map{f(t^1,\ldots,t^n)}$    & := & $\hol{(f \map{t^1} \ldots \map{t^n})}$  \quad
              & for all $n$-ary $f\in F$\\
 & & &  $\hol{f}$ has type $\hol{\underbrace{{i \ar
               \ldots  \ar i} \ar }_{n\geq o} i}$\\
%$\map{p}$    & := & $\hol{p_o}$ & for all $0$-ary $p\in \P$\\
$\map{s = t}$    & := & $\hol{\map{s} = \map{t}}$ \\
$\map{p(t^1,\ldots,t^n)}$    & := & $\hol{(p \map{t^1} \ldots \map{t^n})}$  \quad
             &  for all $n$-ary $p\in P$ \\ 
   & & & $\hol{p}$ has type $\hol{\underbrace{{i \ar
               \ldots  \ar i} \ar }_{n\geq o} o}$\\
$\map{\neg \varphi}$ & := &  $\hol{\neg\map{\varphi}}$  \\
$\map{\varphi \Impl \psi}$ & := &  $\hol{ \map{\varphi} \Impl \map{\psi}}$  \\
$\map{\forall x. s}$ & := &  $\hol{\forall X_i. \ex
                       X_i \Impl \map{s}}$   \\
$\map{\Descr x. s}$ & := &  $\hol{\textbf{IfThenElse}}$ \\
  & & \quad $\hol{(\exists x. \ex x \wedge \map{s} \wedge (\forall
      y. (\ex y \wedge ((\lambda X_i. \map{s}) y)) \Impl y = x))}$ \\
  & & \quad $\hol{ (\Descr  X_i. \map{s})}$ \\
  & & \quad $\hol{\star}$  \\
\end{tabular}
where $\hol{\textbf{IfThenElse}}$ is an abbreviation for the term
$\hol{\lam{S_o}\lam{X_i}\lam{Y_i}. \Descr Z_i. (S_o \wedge Z = X) \vee
(\neg S_o \wedge Z
= Y)}$.

Further connectives can be introduced as usual: $r \vee  s := \neg r
\Impl s$, $r \wedge  s := \neg (\neg r \vee \neg s)$, 
\end{definition}

Todo ... the embedding in Isabelle/HOL is depicted in 
figure \ref{pic1}

% In this framework partial and total functions are modelled as follows: 
% A function $f$ is total if and only if for all $x$ we have $\ex x \imp \ex(f x)$ 
% For partial functions $f$ we may have some $x$ such that $\ex x$ but not
%  $\ex(f x)$. A function $f$ is strict  if  and only if for all $x$
% we have $\ex(f x) \imp \ex x$.

@@

The stepwise evolution has been described in \cite{R58}. Below we
summarise these experiments.  However, first we describe some basic
modeling decisions for the technical encoding in Isabelle/HOL. The
Isabelle/HOL sources of our experiments are available at
\url{todo-sources}. Note that in these sources we did add neither the
definite description operator nor the designated undefined object
$\star$, since both were not required in our experiments.
